Understanding the Landscape of Childlessness in the U.S.: A Data-Driven Exploration (1972-2022)

### 2024-03-11 outputs/paper/paper.qmd
Fig's labs font size changed and some Typo fixed

This paper aims to unpack these factors and explore their impact on the rise of childlessness in America. By analyzing the data, we seek to answer crucial questions: What are the key factors contributing to this trend? How does childlessness manifest differently across demographic groups? Finally, what are the potential social, economic, and individual consequences of this new family structure in the United States? Understanding these dynamics is crucial in having an informed policy discussions and navigating the implications of childlessness in a rapidly changing world.

@@

* Children: This variable captures the number of children ever born alive to the respondent, providing a direct measure of childlessness (having no children).
* Year: This variable allows to analyze trends in childlessness over time, encompassing the chosen time frame for the survey.
* Age: This variable, categorized into relevant age groups (e.g., 18-34, 35-49, etc.), enables us to understand how childlessness manifests across different age demographics.
It is important to note that the GSS data underwent specific naming procedures for this analysis. For instance, the original "CHILD" variable was renamed into the the "childs" variable used here, and age categories were grouped for visualization purposes.

@@

* Socioeconomic Class: Data on socioeconomic class, often measured by income or occupation, will be utilized to understand the impact of economic factors on childbearing choices.
* Educational Attainment: Data on educational attainment, such as highest degree obtained, will be incorporated to explore the relationship between educational level and childbearing. This data can be obtained from sources like the National Center for Education Statistics.

@tbl-nochild-class reveals a potential link between socioeconomic class and childlessness. The data suggests that individuals in the lower class and working class exhibit higher percentages of childlessness compared to those in the middle class and upper class. This pattern might be attributed to various factors, such as financial constraints, job insecurity, or limited access to childcare resources experienced by individuals in lower socioeconomic classes.

@@

* Health Conditions: This variable, measured by medical check-ups of individuals who have been surveyed, will be used to clarify the the context behind childlessness of those with varying health status.
* Marital Status: This variable enables us to see if marriage is a determining factor in childbearing decisions.

@tbl-nochild-health uncovers the relationship between health conditions and childlessness. Intriguingly, those with 'excellent' and 'good' health are substantially more likely to be childless than those with 'fair' and 'poor' health. Several factors are attributable to this fact, but foremost careeer, lifestyle differences and financial considerations might come into play. Though, more data is needed to clarify the context of this trend, especially coupling it with variables like educational attainment or socioeconomic class. In addition, sampling bias might factor into this trend, where it's assumable that people with excellent health are often the younger population and those with less desirable health are older.

@@

ggplot(nochild_age_data_long, aes(x = as.integer(year), y = percentage, color = Age)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE, linetype = "solid", size = 1.2) +  # Add running average trend line
  labs(x = "Year", y = "Percentage of Population", title = "Childless Percentage of Population over Time by Age Group", color = "Age Group") +
  theme_minimal()

@@

It is important to acknowledge, however, that the data series has gaps for several years within this timeframe (1979, 1981, 1992, 1995, 1997, 1999, 2001, 2003, 2005, 2007, 2009, 2011, 2013, 2015, 2017, and 2019). These missing data points can potentially limit the generalization of the observed trends and hinder a comprehensive understanding of the childlessness over time.
Despite these limitations, the available data reveals a steady increase in the percentage of individuals without children across all age groups. Notably, the most significant rise is observed among younger adults (18-34 and 35-49). As an example, the percentage of childless individuals in the 18-34 age group increased by approximately 42.70% from 1972 to 2022.

@fig-nochild-class depicts the gradual increase in the percentage of childless individuals across four social classes (Lower Class, Working Class, Middle Class, Upper Class) in the United States from 1972 to 2022. The graph reveals an upward trend in childlessness for all classes over the five decades.

@@

# Plot the data
ggplot(data_long, aes(x = Year, y = Percentage, color = `Class Identification`)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE, linetype = "solid", size = 1.2) +  # Add running average trend line
  labs(x = "Year", y = "Percentage of Population", title = "Childless Percentage of Population over Time by Class", color = "Breakdown Category") +
  theme_minimal()

@@

@fig-nochild-degree presents a comparative analysis of childlessness across individuals with different levels of educational attainment (Less than High School, High School and College's Degree or Higher) in the United States from 1972 to 2022. The graph utilizes line trends to depict the percentage of childless individuals within each degree category over the five decades.

The data reveals a distinct pattern across educational groups. Individuals with a high school diploma consistently exhibit the highest percentage of childlessness throughout the entire period, ranging from around 28.15% in 1972 to 34.56% in 2022. Conversely, The percentage of childless individuals with a college and higher degree has decreased from 32.85% in 1972 to 29.84% in 2022. The trend for individuals with a with less than a high school education increased from 16.97% in 1972 to 22.90% in 2022.
```{r}
#| echo: false
#| include: true
#| warning: false
#| label: fig-nochild-degree
#| fig-cap: "Percentage Childless over Time by Level of Education"

@@

# Plot the data
ggplot(data_long, aes(x = Year, y = Percentage, color = `Degree`)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed", size = 1.2) +  # Add running average trend line
  labs(x = "Year", y = "Percentage of Population", title = "Childless Percentage of Population over Time by Degree", color = "Degree") +
  theme_minimal()

@@

# Plot the data
ggplot(data_long, aes(x = Year, y = Percentage, color = `Gender`)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed", size = 1.2) +  # Add running average trend line
  labs(x = "Year", y = "Percentage of Population", title = "Childless Percentage of Population over Time by Gender", color = "Gender") +
  theme_minimal()

@@

@fig-nochild-health explores health differences in childlessness within the United States over a 50-year period (1972-2022). The graph utilizes trend lines to depict the percentage of childless individuals categorized by health conditions (Excellent, Good, Fair Poor).

Overall, the data shows an increase in childlessness across all health levels and conditions approaching 2022. As mentioned, people with excellent health has the highest percentage of childlessness, having a 10.49% increase throughout the 50-year period (28.26% in 1972 to 38.75% in 2022). People with good and fair health exhibit similar rate of childlessness in 2022, at about 30%. In particular, people with fair health markedHowever, people with poor health, once having the lowest rate of childlessness (at 12.27% in 1972), now increases to 20.40% in 2022, which translates to a substantial 8.13% increase.

```{r}
#| echo: false
#| include: true
#| warning: false
#| label: fig-nochild-health
#| fig-cap: "Percentage Childless over Time by Health Conditions"

@@

# Plot the data
ggplot(data_long, aes(x = Year, y = Percentage, color = `Condition of health`)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed", size = 1.2) +  # Add running average trend line
  labs(x = "Year", y = "Percentage of Population", title = "Childless Percentage of Population over Time by Health", color = "Condition of health") +
  theme_minimal()

@@

@fig-nochild-marital explores marital status of people who are childless over a span of 50 years in the United States (1972-2022). The graph contains trend lines of varying marital statuses-Divorced, Married, Never married, Separated and Widowed-and how each has changed over the years in terms of childbearing status.

The data demonstrates that a significant portion of people who have never married are childless. However, more and more people are reluctant to get married but still consider having children. This is observable in the childlessness percentage decrease of people who never marry, decreasing from 96.73% in 1972 to 72.53% in 2022. For people who are divorced, separated and widowed, this percentage fluctuates, displaying the complex dynamics between these relationship status and childbearing. For married people though, this childlessness percentage also decreases, from 10.51% in 1972 to 13.97% in 2022.

@@

# Plot the data
ggplot(data_long, aes(x = Year, y = Percentage, color = `Marital status`)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed", size = 1.2) +  # Add running average trend line
  labs(x = "Year", y = "Percentage of Population", title = "Childless Percentage of Population over Time by Marital Status", color = "Marital status") +
  theme_minimal()

@@

The figures presented in this paper were generated using data from the [@gss]. These figures visually represent the key findings related to the trends and patterns in childlessness across different age, social classes, educational attainment levels, and genders in the United States over the past five decades (1972-2022). It is important to remember that these figures solely present descriptive data and do not establish causal relationships. The Discussion section will further analyze the observed trends and explore the broader implications of these findings.

@@

To gain a deeper understanding of childlessness, there are some crucial next steps that need to be taken. Firstly, employing longitudinal studies will allow the exploration of causal relationships between factors and childbearing decisions over time. Secondly, incorporating diverse data sources, such as qualitative interviews and other surveys with broader representation, can provide richer insights and explore the nuances of individual experiences. Lastly, utilizing a mixed methods approach, combining quantitative and qualitative data analysis, will offer a more comprehensive understanding of both the broader trends and the lived experiences of individuals facing childlessness.

>>>>

This paper aims to unpack these factors and explore their impact on the rise of childlessness in America. By analyzing the data, we seek to answer crucial questions: What are some important factors contributing to this trend? How does childlessness manifest differently across demographic groups? Finally, what are the potential social, economic, and individual consequences of this new family structure in the United States? Understanding these dynamics is crucial in having an informed policy discussion and navigating the implications of childlessness in a rapidly changing world.

@@

* Children: This variable captures the number of children ever born alive to the respondent, providing a direct measure of childlessness (having no children).
* Year: This variable allows us to analyze trends in childlessness over time, encompassing the chosen time frame for the survey.
* Age: This variable, categorized into relevant age groups (e.g., 18-34, 35-49, etc.), enables us to understand how childlessness manifests across different age demographics.

@@

* Socioeconomic Class: Data on socioeconomic class, often measured by income or occupation, will be utilized to understand the impact of economic factors on childbearing choices.
* Educational Attainment: Data on educational attainment, such as the highest degree obtained, will be incorporated to explore the relationship between educational level and childbearing.

@tbl-nochild-class reveals a potential link between socioeconomic class and childlessness. The data suggests that individuals in the lower and working classes exhibit higher percentages of childlessness compared to those in the middle and upper classes. This pattern might be attributed to various factors, such as financial constraints, job insecurity, or limited access to childcare resources experienced by individuals in lower socioeconomic classes.

@@

From the GSS, we utilize several important variables relevant to our analysis:

* Health Conditions: This variable, measured by medical check-ups of individuals who have been surveyed, will be used to clarify the context behind the childlessness of those with varying health statuses.
* Marital Status: This variable enables us to see if marriage is a determining factor in childbearing decisions.

@tbl-nochild-health uncovers the relationship between health conditions and childlessness. Intriguingly, those with 'excellent' and 'good' health are substantially more likely to be childless than those with 'fair' and 'poor' health. Several factors are attributable to this fact, but foremost, career, lifestyle differences and financial considerations might come into play. However, more data is needed to clarify the context of this trend, especially coupling it with variables like educational attainment or socioeconomic class. In addition, sampling bias might factor into this trend, where it's assumable that people with excellent health are often the younger population and those with less desirable health are older.

@@

ggplot(nochild_age_data_long, aes(x = as.integer(year), y = percentage, color = Age)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed", size = 1.2) +  # Add running average trend line
  labs(x = "Year", y = "Percentage of Population", title = "Childless Percentage of Population over Time by Age Group", color = "Age Group") +
  theme_minimal() +
  theme(text = element_text(size = 10))

@@

It is important to acknowledge, however, that the data series has gaps for several years within this timeframe (1979, 1981, 1992, 1995, 1997, 1999, 2001, 2003, 2005, 2007, 2009, 2011, 2013, 2015, 2017, and 2019). These missing data points can potentially limit the generalization of the observed trends and hinder a comprehensive understanding of childlessness over time.
Despite these limitations, the available data reveals a steady increase in the percentage of individuals without children across all age groups. Notably, the most significant rise is observed among younger adults (18-34 and 35-49). As an example, the percentage of childless individuals in the 18-34 age group increased by approximately 42.70% from 1972 to 2022.

@fig-nochild-class depicts the gradual increase in the percentage of childless individuals across four social classes (lower class, working class, middle class, and upper class) in the United States from 1972 to 2022. The graph reveals an upward trend in childlessness for all classes over the five decades.

@@

# Plot the data
ggplot(data_long, aes(x = Year, y = Percentage, color = `Class Identification`)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed", size = 1.2) +  # Add running average trend line
  labs(x = "Year", y = "Percentage of Population", title = "Childless Percentage of Population over Time by Class", color = "Breakdown Category") +
  theme_minimal() +
  theme(text = element_text(size = 10))

@@

@fig-nochild-degree presents a comparative analysis of childlessness across individuals with different levels of educational attainment (less than high school, high school and college degree or higher) in the United States from 1972 to 2022. The graph utilizes line trends to depict the percentage of childless individuals within each degree category over the five decades.

The data reveals a distinct pattern across educational groups. Individuals with a high school diploma consistently exhibit the highest percentage of childlessness throughout the entire period, ranging from around 28.15% in 1972 to 34.56% in 2022. Conversely, the percentage of childless individuals with a college and higher degree has decreased from 32.85% in 1972 to 29.84% in 2022. The trend for individuals with a with less than a high school education increased from 16.97% in 1972 to 22.90% in 2022.
```{r}
#| echo: false
#| include: true
#| warning: false
#| label: fig-nochild-degree
#| fig-cap: "Percentage Childless over Time by Level of Education"

@@

# Plot the data
ggplot(data_long, aes(x = Year, y = Percentage, color = `Degree`)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed", size = 1.2) +  # Add running average trend line
  labs(x = "Year", y = "Percentage of Population", title = "Childless Percentage of Population over Time by Degree", color = "Degree") +
  theme_minimal() +
  theme(text = element_text(size = 10))

@@

# Plot the data
ggplot(data_long, aes(x = Year, y = Percentage, color = `Gender`)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed", size = 1.2) +  # Add running average trend line
  labs(x = "Year", y = "Percentage of Population", title = "Childless Percentage of Population over Time by Gender", color = "Gender") +
  theme_minimal() +
  theme(text = element_text(size = 10))

@@

@fig-nochild-health explores health differences in childlessness within the United States over a 50-year period (1972-2022). The graph utilizes trend lines to depict the percentage of childless individuals categorized by health conditions (excellent, good, fair, and poor).

Overall, the data shows an increase in childlessness across all health levels and conditions approaching 2022. As mentioned, people with excellent health have the highest percentage of childlessness, having a 10.49% increase throughout the 50-year period (28.26% in 1972 to 38.75% in 2022). People with good and fair health exhibit a similar rate of childlessness in 2022, at about 30%. In particular, people with fair health marked; however, people with poor health, once having the lowest rate of childlessness (at 12.27% in 1972), now increases to 20.40% in 2022, which translates to a substantial 8.13% increase.  
```{r}
#| echo: false
#| include: true
#| warning: false
#| label: fig-nochild-health
#| fig-cap: "Percentage Childless over Time by Health Conditions"

@@

# Plot the data
ggplot(data_long, aes(x = Year, y = Percentage, color = `Condition of health`)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed", size = 1.2) +  # Add running average trend line
  labs(x = "Year", y = "Percentage of Population", title = "Childless Percentage of Population over Time by Health", color = "Condition of health") +
  theme_minimal() +
  theme(text = element_text(size = 10))

@@

@fig-nochild-marital explores the marital status of people who are childless over a span of 50 years in the United States (1972-2022). The graph contains trend lines of varying marital statuses-Divorced, Married, Never married, Separated and Widowed-and how each has changed over the years in terms of childbearing status.

The data demonstrates that a significant portion of people who have never married are childless. However, more and more people are reluctant to get married but still consider having children. This is observable in the childlessness percentage decrease of people who never marry, decreasing from 96.73% in 1972 to 72.53% in 2022. For people who are divorced, separated and widowed, this percentage fluctuates, displaying the complex dynamics between these relationship statuses and childbearing. For married people though, this childlessness percentage also decreases, from 10.51% in 1972 to 13.97% in 2022.

@@

# Plot the data
ggplot(data_long, aes(x = Year, y = Percentage, color = `Marital status`)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed", size = 1.2) +  # Add running average trend line
  labs(x = "Year", y = "Percentage of Population", title = "Childless Percentage of Population over Time by Marital Status", color = "Marital status") +
  theme_minimal() +
  theme(text = element_text(size = 10))

@@

The figures presented in this paper were generated using data from the [@gss]. These figures visually represent the key findings related to the trends and patterns in childlessness across different ages, social classes, educational attainment levels, and genders in the United States over the past five decades (1972-2022). It is important to remember that these figures solely present descriptive data and do not establish causal relationships. The Discussion section will further analyze the observed trends and explore the broader implications of these findings.

@@

To gain a deeper understanding of childlessness, some crucial next steps need to be taken. Firstly, employing longitudinal studies will allow the exploration of causal relationships between factors and childbearing decisions over time. Secondly, incorporating diverse data sources, such as qualitative interviews and other surveys with broader representation, can provide richer insights and explore the nuances of individual experiences. Lastly, utilizing a mixed methods approach, combining quantitative and qualitative data analysis, will offer a more comprehensive understanding of both the broader trends and the lived experiences of individuals facing childlessness.
